Follow-up Samples

### 2017-08-12 submission/manuscript_R1.Rmd
complete Pat edits upto Results

***Treatment for colonic lesions alters the bacterial community structure***
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), or carcinoma (N = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1]. None of the individuals in this study had any recorded antibiotic usage that was not associated with surgical treatment of their respective lesion. The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A]. We found that the communities obtained pre and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001). There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05). Since surgery is a procedure where antibiotics are used, we tested our $\theta$~YC~ beta diversity in these individuals only. Although not significant the trends persisted (P-value `r ifelse(min(surg_only_comparisons[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1A]. There was also no difference in the change of $\theta$~YC~ between pre and post treatment samples between those who had surgery versus those who did not for both the adenoma and advanced adenoma groups (P-value `r ifelse(min(surgVnonsurg_noncrc_pvalues[, "BH"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure S1B-C]. Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D]. We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value = `r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value = `r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value = `r beta_summary["Carcinoma", "PERMANOVA"]`). Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value > 0.05) [Table S1].

***The treatment of lesions are not consistent across diagnosis groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) > 0.05, invisible(">"), invisible("<"))` 0.05). It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test. We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. To try and limit the likelihood that the models would overfit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs. Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. Although treatment had an impact on the overall community structure; the effect of treatment was not consistent across patients and diagnosis groups.

***Post-treatment samples from patients with carcinoma more closely resemble those of a normal colon.***
Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons. To test this, we used an expanded cohort of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma [Table 2]. We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon. The models performed well (adenoma AUC range =`r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2). The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A]. Among the OTUs that were shared across the three models were those populations generally considered beneficial to their host (e.g. *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B]. Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S3] with some individuals showing a marked decrease in relative abundance [Figure S4]. Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model. A decrease in the positive probability would indicate that the microbiota more closely resembled that of a patient with a normal colon. There was no significant change in the positive probability for the adenoma or advanced adenoma groups (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] > 0.05 & all_probs_paired_wilcox["SRN", "BH_correction"] > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4]. The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals (P-value = `r format(all_probs_paired_wilcox["carcinoma", "BH_correction"], digits = 3)`). Only, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) who were diagnosed with a carcinoma had a higher positive probability after treatment; one of those was re-diagnosed with carcinoma on the follow up visit. These results indicate that, although there were changes in the microbiota associated with treatment, those experienced by  patients with carcinoma after treatment yielded gut bacterial communities of greater similarity to that of a normal colon.

***Difficult to identify effects of specific treatments on the change in the microbiota.***
The type of treatment that the patients received varied across diagnosis groups. Those with adenomas and advanced adenomas received surgical resection (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "Y")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "Y")))`) or polyp removal during colonoscopy (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "N")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "N")))`) and those with carcinomas received surgical resection (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "no")))`), surgical resection with chemotherapy (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "no")))`), and surgical resection with chemotherapy and radiation (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "yes")))`). Regardless of treatment used there was no significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Furthermore, there was no significant difference in the effect of the treatments on the amount of change in the community structure (P-value = `r ifelse(chemo_rad_summary["dist", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`). Finally, the change in the positive probability was also not significantly different between any treatment group (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`). Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota.

>>>>

***Treatment for colonic lesions alters the bacterial community structure***
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), or carcinoma (N = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1]. None of the individuals in this study had any recorded antibiotic usage that was not associated with surgical treatment of their respective lesion. The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A]. We found that the communities obtained pre and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001). There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05). Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D]. We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value = `r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value = `r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value = `r beta_summary["Carcinoma", "PERMANOVA"]`). Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value > 0.05) [Table S1].

***The treatment of lesions are not consistent across diagnosis groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) > 0.05, invisible(">"), invisible("<"))` 0.05). It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test. We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. To limit the likelihood that the models would overfit the data because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs. Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. Although treatment had an impact on the overall community structure, the effect of treatment was not consistent across patients and diagnosis groups.

***Post-treatment samples from patients with carcinoma more closely resemble those of a normal colon.***
Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons. To test this, we used an expanded cohort of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma [Table 2]. We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon. The models performed well (adenoma AUC range = `r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S2). The OTUs that were incorporated into the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A]. Among the OTUs that were shared across the three models were those populations generally considered beneficial to their host (e.g. *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B]. Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S3] with some individuals showing a marked decrease in relative abundance [Figure S4]. Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model. A decrease in the positive probability would indicate that the microbiota more closely resembled that of a patient with a normal colon. There was no significant change in the positive probability for the adenoma or advanced adenoma groups (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] > 0.05 & all_probs_paired_wilcox["SRN", "BH_correction"] > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4]. The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals (P-value = `r format(round(all_probs_paired_wilcox["carcinoma", "BH_correction"], digits = 2))`). Only, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) who were diagnosed with a carcinoma had a higher positive probability after treatment; one of those was re-diagnosed with carcinoma on the follow up visit. These results indicate that, although there were changes in the microbiota associated with treatment, those experienced by  patients with carcinoma after treatment yielded gut bacterial communities of greater similarity to that of a normal colon.

***Difficult to identify effects of specific treatments on the change in the microbiota.***
The type of treatment that the patients received varied across diagnosis groups. Those with adenomas and advanced adenomas received surgical resection (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "Y")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "Y")))`) or polyp removal during colonoscopy (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "N")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "N")))`) and those with carcinomas received surgical resection (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "no")))`), surgical resection with chemotherapy (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "no")))`), and surgical resection with chemotherapy and radiation (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "yes")))`). Regardless of treatment used there was no significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P-value = `r ifelse(chemo_rad_summary["dist", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`). Finally, the change in the positive probability was not significantly different between any of the treatment groups (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`). Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota.
